---
title: "105-not-erebi-sl"
output: html_document
date: "2025-12-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r}
library(ape)
library(vcfR)
library(adegenet)
library(ggpubr)
library(ggrepel)
library(ggplotify)
library(snpR)
library(ggtree)
library(tanggle)
library(RColorBrewer)
library(pheatmap)
library(tidyverse)
```

Received a VCF from Luis on 12/2

Have list of chroms in data/chroms.tsv


scaffold_6 6291620	22526456
scaffold_8 5067926	29962161
scaffold_9 14123901	35865842
scaffold_11 219668	24099434
scaffold_21 6558641	24209072

Looks to have 80,783 variants and match Peter's sample names
These have mostly been filtered for missing data but can check 


```{r}
meta<-read_csv("meta/meta-filtered-edited.csv")
meta$Taxon<-factor(meta$Taxon, levels=c("erebi","richardsoni","bulleri","paracome","Pilbara sp. nov.","papuensis","come","vlaminghi"))
meta<-meta %>% mutate(Label=paste0(pop,"_",id))
```

# Not erebi

We'll look at 'erebi' excluding erebi

```{r}
noterebi<-meta %>% filter(!Taxon %in%c("erebi","papuensis","come","vlaminghi"))
noterebi %>% select(Label) %>% write_tsv(file = "meta/noterebi.tsv", col_names = FALSE)
```

```{sh, eval=FALSE}
bcftools view -S meta/noterebi.tsv data/gl_vcf.vcf.gz > outputs/105/noterebi.vcf
bcftools +fill-tags outputs/105/noterebi.vcf  -- -t MAF,F_MISSING | bcftools view -i 'F_MISSING < 0.05' | bcftools view -i 'INFO/MAF > 0.05'  > outputs/105/noterebi-filtered.vcf
bgzip outputs/105/noterebi-filtered.vcf
tabix outputs/105/noterebi-filtered.vcf.gz

mkdir outputs/105/vcfs
mkdir outputs/105/chrom-ld
mkdir outputs/105/chrom-pcs
```

```{sh, eval=FALSE}
paste \
<(bcftools query -f '[%SAMPLE\t]\n' noterebi-filtered.vcf.gz | head -1 | tr '\t' '\n') \
<(bcftools query -f '[%GT\t]\n' noterebi-filtered.vcf.gz | awk -v OFS="\t" '{for (i=1;i<=NF;i++) if ($i == "./.") sum[i]+=1 } END {for (i in sum) print i, sum[i] / NR }' | sort -k1,1n | cut -f 2) > missingness.stats
 cat missingness.stats | sort -k 2 -n | tail
```

Finniss_NEFinniss.1	0.0519218
Dillie_NEDillie.1	0.0520904
Dillie_NEDillie.2	0.0520904
Ashburton_NEAshburton.1	0.0591706
Blina_NEBlina.1	0.0647336
Clints_NEClints.1	0.0686109
Liverpool_NELiverpool.1	0.0770398
Manning_NEManning.2	0.0957519
Manning_NEManning.1	0.0962576
Koolatong_NEKoolatong.1	0.152394

## LD
```{sh, eval=FALSE}
cat data/chroms.tsv  | while read line; do bcftools view -Ov -r $line outputs/105/noterebi-filtered.vcf.gz > outputs/105/vcfs/$line.vcf; done;

bcftools view  outputs/105/noterebi-filtered.vcf.gz -r scaffold_6:6291620-22526456 > outputs/105/vcfs/6-region.vcf
bcftools view  outputs/105/noterebi-filtered.vcf.gz -r scaffold_8:5067926-29962161 > outputs/105/vcfs/8-region.vcf
bcftools view  outputs/105/noterebi-filtered.vcf.gz -r scaffold_9:14123901-35865842 > outputs/105/vcfs/9-region.vcf
bcftools view  outputs/105/noterebi-filtered.vcf.gz -r scaffold_11:219668-24099434 > outputs/105/vcfs/11-region.vcf
bcftools view  outputs/105/noterebi-filtered.vcf.gz -r scaffold_21:6558641-24209072 > outputs/105/vcfs/21-region.vcf

# in outputs/105/vcfs
for f in *.vcf; do plink --vcf $f --r2 inter-chr --ld-window-r2 0.1 --out `basename $f vcf`ldf --allow-extra-chr --double-id; done;

```

```{r, eval=FALSE}
files<-list.files("outputs/105/vcfs",pattern = "*.ldf.ld", full.names = TRUE)

plotLd<-function(file) {
  chrom<-gsub("outputs/105/vcfs/","",file)
  chrom<-gsub(".ldf.ld","", chrom)
  lc<-read.delim(file,sep="",stringsAsFactors=FALSE) %>% as_tibble() %>% arrange(R2) %>%  filter(R2 >0.3)

  ggplot(lc) +
  geom_point(aes(x=BP_A, y=BP_B, color=R2), alpha=0.5) +
  scale_color_gradient(low="khaki1", high="red") + 
  ggtitle(paste0(chrom))+
  theme_bw() +
  theme(panel.grid = element_blank())
ggsave(paste0("outputs/105/chrom-ld/",chrom,".pdf"))
}

lapply(files, plotLd)
```

## Chrom by Chrom PCs


```{r, eval=FALSE}
files<-list.files("outputs/105/vcfs",pattern = "*.vcf", full.names = TRUE)

plotPCA<-function(file) {
  meta<-noterebi
  chrom<-gsub("outputs/105/vcfs/","",file)
  chrom<-gsub(".vcf","", chrom)
  
  vcf<-read.vcfR(file=file)
  genind<-vcfR2genind(vcf)
  genind@pop<-as.factor(meta$Pop)

gen<-genind
X <- tab(gen, NA.method="mean")

pca1 <- dudi.pca(X,scannf=FALSE,scale=FALSE,nf=3)

df<-pca1$li
df<-df %>% bind_cols(meta)
eig<-pca1$eig/sum(pca1$eig)*100


pc12<-ggplot(df) + 
  geom_point(aes(x=Axis1, y=Axis2, fill=Taxon),pch=21, alpha=0.75, cex=2) +
  theme_bw() +
  xlab(paste0("PC1 ",round(eig[1],2), "%")) +
  ylab(paste0("PC2 ",round(eig[2],2), "%")) +
  theme(panel.grid = element_blank()) +
  scale_fill_viridis_d(option="H") +
  guides(fill = guide_legend(override.aes = list(pch = 22))) +
  theme(legend.position = "top") 
  #geom_text_repel(aes(x=Axis1, y=Axis2,label=ID))
pc12

ggsave(paste0("outputs/105/chrom-pcs/",chrom,".pdf"))
}

lapply(files, plotPCA)
```

Identify genotypes here.

# Scaffold 6


```{r}
vcf<-read.vcfR(file="outputs/105/vcfs/6-region.vcf")
genind<-vcfR2genind(vcf)
genind@pop<-as.factor(noterebi$Taxon)
gen<-genind
X <- tab(gen, NA.method="mean")

pca1 <- dudi.pca(X,scannf=FALSE,scale=FALSE,nf=3)

df<-pca1$li
df<-df %>% bind_cols(noterebi)

eig<-pca1$eig/sum(pca1$eig)*100
```

```{r}
pc12<-ggplot(df) + 
  geom_point(aes(x=Axis1, y=Axis2, fill=Taxon), alpha=0.75, cex=2, pch=21) +
  theme_bw() +
  xlab(paste0("PC1 ",round(eig[1],2), "%")) +
  ylab(paste0("PC2 ",round(eig[2],2), "%")) +
  geom_vline(xintercept = -10) +
  geom_vline(xintercept = -2) +
  theme(panel.grid = element_blank()) +
  scale_fill_viridis_d(option="H") +
  guides(fill = guide_legend(override.aes = list(pch = 22))) +
  theme(legend.position = "top") 
pc12
ggsave("outputs/105/scaffold6-region-pca.jpeg")
```

```{r}
df %>% filter(Axis2 < -10)
df %>% filter(Axis2 >-10 & Axis2 < -2)

df2<-df %>% mutate(Chromosome="6") %>% mutate(Genotype=ifelse(Axis1 < -10, "RHom",ifelse(Axis1 > -10 & Axis1 < -2,"Het","AHom"))) %>%
  select(Label,Genotype,Chromosome)
write_csv(df2,"meta/noterebi-ss-chrom6-genos.csv")

df2%>% group_by(Genotype) %>% summarize(Count=n())
```


# Scaffold 8


```{r}
vcf<-read.vcfR(file="outputs/105/vcfs/8-region.vcf")
genind<-vcfR2genind(vcf)
genind@pop<-as.factor(noterebi$Taxon)
gen<-genind
X <- tab(gen, NA.method="mean")

pca1 <- dudi.pca(X,scannf=FALSE,scale=FALSE,nf=3)

df<-pca1$li
df<-df %>% bind_cols(noterebi)

eig<-pca1$eig/sum(pca1$eig)*100
```

```{r}
pc12<-ggplot(df) + 
  geom_point(aes(x=Axis1, y=Axis2, fill=Taxon), alpha=0.75, cex=2, pch=21) +
  theme_bw() +
  xlab(paste0("PC1 ",round(eig[1],2), "%")) +
  ylab(paste0("PC2 ",round(eig[2],2), "%")) +
  geom_vline(xintercept = -10) +
  geom_vline(xintercept = 2) +
  theme(panel.grid = element_blank()) +
  scale_fill_viridis_d(option="H") +
  guides(fill = guide_legend(override.aes = list(pch = 22))) +
  theme(legend.position = "top") 
pc12
ggsave("outputs/105/scaffold8-region-pca.jpeg")
```

```{r}
df2<-df %>% mutate(Chromosome="8") %>% mutate(Genotype=ifelse(Axis1 < -10, "RHom",ifelse(Axis1 > -10 & Axis1 < 2,"Het","AHom"))) %>%
  select(Label,Genotype,Chromosome)
write_csv(df2,"meta/noterebi-ss-chrom8-genos.csv")

df2%>% group_by(Genotype) %>% summarize(Count=n())
```


# Scaffold 9


```{r}
vcf<-read.vcfR(file="outputs/105/vcfs/9-region.vcf")
genind<-vcfR2genind(vcf)
genind@pop<-as.factor(noterebi$Taxon)
gen<-genind
X <- tab(gen, NA.method="mean")

pca1 <- dudi.pca(X,scannf=FALSE,scale=FALSE,nf=3)

df<-pca1$li
df<-df %>% bind_cols(noterebi)

eig<-pca1$eig/sum(pca1$eig)*100
```

```{r}
pc12<-ggplot(df) + 
  geom_point(aes(x=Axis1, y=Axis2, fill=Taxon), alpha=0.75, cex=2, pch=21) +
  theme_bw() +
  xlab(paste0("PC1 ",round(eig[1],2), "%")) +
  ylab(paste0("PC2 ",round(eig[2],2), "%")) +
  geom_vline(xintercept = -10) +
  geom_vline(xintercept = 2) +
  theme(panel.grid = element_blank()) +
  scale_fill_viridis_d(option="H") +
  guides(fill = guide_legend(override.aes = list(pch = 22))) +
  theme(legend.position = "top") 
pc12
ggsave("outputs/105/scaffold9-region-pca.jpeg")
```

```{r}

df2<-df %>% mutate(Chromosome="9") %>% mutate(Genotype=ifelse(Axis1 < -10, "RHom",ifelse(Axis1 > -10 & Axis1 < 2,"Het","AHom"))) %>%
  select(Label,Genotype,Chromosome)
write_csv(df2,"meta/noterebi-ss-chrom9-genos.csv")

df2%>% group_by(Genotype) %>% summarize(Count=n())
```



# Scaffold 11

```{r}
vcf<-read.vcfR(file="outputs/105/vcfs/11-region.vcf")
genind<-vcfR2genind(vcf)
genind@pop<-as.factor(noterebi$Taxon)
gen<-genind
X <- tab(gen, NA.method="mean")

pca1 <- dudi.pca(X,scannf=FALSE,scale=FALSE,nf=3)

df<-pca1$li
df<-df %>% bind_cols(noterebi)

eig<-pca1$eig/sum(pca1$eig)*100
```

```{r}
pc12<-ggplot(df) + 
  geom_point(aes(x=Axis1, y=Axis2, fill=Taxon), alpha=0.75, cex=2, pch=21) +
  theme_bw() +
  xlab(paste0("PC1 ",round(eig[1],2), "%")) +
  ylab(paste0("PC2 ",round(eig[2],2), "%")) +
  geom_vline(xintercept = -7) +
  geom_vline(xintercept = 6) +
  theme(panel.grid = element_blank()) +
  scale_fill_viridis_d(option="H") +
  guides(fill = guide_legend(override.aes = list(pch = 22))) +
  theme(legend.position = "top") 
pc12
ggsave("outputs/105/scaffold11-region-pca.jpeg")
```

```{r}
df2<-df %>% mutate(Chromosome="11") %>% mutate(Genotype=ifelse(Axis1 < -7, "RHom",ifelse(Axis1 > -7 & Axis1 < 6,"Het","AHom"))) %>%
  select(Label,Genotype,Chromosome)
write_csv(df2,"meta/noterebi-ss-chrom11-genos.csv")

df2%>% group_by(Genotype) %>% summarize(Count=n())
```

# Scaffold 21

```{r}
vcf<-read.vcfR(file="outputs/105/vcfs/21-region.vcf")
genind<-vcfR2genind(vcf)
genind@pop<-as.factor(noterebi$Taxon)
gen<-genind
X <- tab(gen, NA.method="mean")

pca1 <- dudi.pca(X,scannf=FALSE,scale=FALSE,nf=3)

df<-pca1$li
df<-df %>% bind_cols(noterebi)

eig<-pca1$eig/sum(pca1$eig)*100
```

```{r}
pc12<-ggplot(df) + 
  geom_point(aes(x=Axis1, y=Axis2, fill=Taxon), alpha=0.75, cex=2, pch=21) +
  theme_bw() +
  xlab(paste0("PC1 ",round(eig[1],2), "%")) +
  ylab(paste0("PC2 ",round(eig[2],2), "%")) +
  geom_vline(xintercept = -5) +
  geom_vline(xintercept = 2) +
  theme(panel.grid = element_blank()) +
  scale_fill_viridis_d(option="H") +
  guides(fill = guide_legend(override.aes = list(pch = 22))) +
  theme(legend.position = "top") 
pc12
ggsave("outputs/105/scaffold21-region-pca.jpeg")
```

```{r}

df2<-df %>% mutate(Chromosome="21") %>% mutate(Genotype=ifelse(Axis1 < -5, "RHom",ifelse(Axis1 > -5 & Axis1 < 2,"Het","AHom"))) %>%
  select(Label,Genotype,Chromosome)
write_csv(df2,"meta/noterebi-ss-chrom21-genos.csv")

df2%>% group_by(Genotype) %>% summarize(Count=n())
```